---
title: "Untitled"
output: html_document
date: "2025-03-24"
---

```{r}
library(survival)
library(haven)
library(dplyr)
```

```{r}
data<-read_sas("/Users/jackielindstrom/Documents/675 Survival/kidney_ecd_1.sas7bdat")
```

```{r}
head(data)
```

a) Fit a model which contains only factors known at the time of transplant (t = 0). List the factors that significantly predict death

```{r}
time<-pmin(data$time_to_death,data$time_to_censor)
head(time)
```


```{r}
#Making delta and Xi  (surv time)
data$status <- ifelse(is.na(data$time_to_death), 0, 1)

data$time <- c(0)

for (i in 1:nrow(data)) {
  if (is.na(data$time_to_death[i])) {
    data$time[i] <- data$time_to_censor[i]  
  } else {
    data$time[i] <- data$time_to_death[i]  
  }
}
```

```{r}
cox_model <- coxph(Surv(time, status==1) ~ age + male + diabetes + comorbid + ECD, data = data)

summary(cox_model)
```
All factors significantly predict time to death because all factors have a p value of less than 0.05. 

b) Interpret the ECD effect from the model from part (a) **ECD=1 for ECD kidney, ECD=0 for non ECD kidney

"ECD kidneys increase death hazard by 13.373%.

(Note: for a cox model, we have hazard=baselinehazard*exp(BZi). So B=log(hazard/basehazard, Note: not 113%)) 

c) Fit a model with graft failure (GF) as a time-dependent covariate (Use document timedep.pdf)

```{r}
# Need to restructure data set with time dep
data$GF_y_n<- ifelse(is.na(data$time_to_GF), 0, 1)
```

```{r}
tdata <- tmerge(data1 = data, 
                data2 = data, 
                id = idnum, 
                tstart = 0, 
                tstop = data$time, 
                death = event(time_to_death, status))

surv_data <- tmerge(data1 = tdata, 
                    data2 = data, 
                    id = idnum, 
                    GFevent = event(time_to_GF), # do they get GF after this point
                    GF = tdc(time_to_GF, GF_y_n)) #up until this point NA until GF

surv_data$GF[is.na(surv_data$GF)] <- 0
```


```{r}
head(surv_data)
```


```{r}
d_cox_model <- coxph(Surv(tstart, tstop, death==1) ~ age + GF + male + diabetes + comorbid + ECD, data = surv_data)
summary(d_cox_model)
```

d) Interpret the hazard ratio for GF
The coefficient is 0.515, and exp(coef)= 1.674.
Individuals with GF have a 67.4% higher chance of dying compared to those without GF. 

e) Compare the ECD hazard ratios from models (a) and (c). What does this tell
you about the nature of the ECD effect?

The ECD ratio is 0.9896 for model (c), but it is not statistically significant. Therefore we could say ECD does not have a significant effect on death after accounting for the graft failure, instead of how in part (a) we said ECD increased risk of death by 13.373%. Adding in the time dependent covariate graft failure GF results in the ECD effect on death hazard decreasing. 

